Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-04-28 01-introduction.qmd
referencias intro y antecedentes

Perception of economic inequality has been associated with important outcomes such as redistributive preferences, justification of inequality, and legitimacy of the economic system (Castillo et al., 2022 \[acá citar gente de Granada\]). Research shows that greater perceived inequality fosters stronger preferences for redistribution, regardless of actual inequality levels [@garcia-sanchez_vicious_2019; @gimpelson_misperceiving_2018]. However, other findings suggest that heightened perceptions of income gaps between high- and low-paying occupations may lead to greater justification of inequality [@Castillo2011]. A less explored area in this regard concerns market justice, which denotes the degree to which individuals consider just that social goods and services (e.g., health, pensions, education) are allocated according to individual contribution, competition, and ability to pay [@lane_market_1986; @lindh_public_2015; @kluegel_legitimation_1999]. Market justice attitudes reflect the belief that the market promotes procedural fairness---equality of opportunity---so that outcomes result from individual achievement [@lane_market_1986]. Indeed, Lindh [-@lindh_public_2015] shows that market justice preferences are stronger in countries with higher private spending on services than in those with more comprehensive welfare systems. From this perspective, perceptions of economic inequality may be key, as lower perceived gaps can reinforce acceptance of free-market systems and justify existing inequalities.

>>>>

Perception of economic inequality has been associated with important outcomes such as redistributive preferences, justification of inequality, and legitimacy of the economic system [@castillo_perception_2022; @garcia-sanchez_perceptions_2018; @garcia-sanchez_attitudes_2020]. Research shows that greater perceived inequality fosters stronger preferences for redistribution, regardless of actual inequality levels [@garcia-sanchez_vicious_2019; @gimpelson_misperceiving_2018]. However, other findings suggest that heightened perceptions of income gaps between high- and low-paying occupations may lead to greater justification of inequality [@Castillo2011]. A less explored area in this regard concerns market justice, which denotes the degree to which individuals consider just that social goods and services (e.g., health, pensions, education) are allocated according to individual contribution, competition, and ability to pay [@lane_market_1986; @lindh_public_2015; @kluegel_legitimation_1999]. Market justice attitudes reflect the belief that the market promotes procedural fairness---equality of opportunity---so that outcomes result from individual achievement [@lane_market_1986]. Indeed, Lindh [-@lindh_public_2015] shows that market justice preferences are stronger in countries with higher private spending on services than in those with more comprehensive welfare systems. From this perspective, perceptions of economic inequality may be key, as lower perceived gaps can reinforce acceptance of free-market systems and justify existing inequalities.

### 2025-04-28 02-antecedents.qmd
referencias intro y antecedentes

Empirical studies have developed different strategies to capture and operationalize market justice preferences. A common approach in the literature is to gauge attitudes toward the legitimacy of inequality in specific domains, especially when linked to income differences. Kluegel and Smith [-@kluegel_beliefs_1981] and later Kluegel et al. [-@kluegel_legitimation_1999] laid the groundwork for measuring beliefs in market justice through survey items that assess whether people support economic inequality when it reflects personal effort and skills. Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, and/or pensions. For example, Knesebeck et al. (2016) and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia. These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts. Some research, like that of Lindh [-@lindh_public_2015] and Svallfors [-@svallfors_political_2007], uses these indicators in combination to measure broader preferences for market-based distribution in welfare systems. More recently, Castillo et al. [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile. These different operationalizations all aim to capture the extent to which individuals accept inequality when it is framed as a reflection of market outcomes.

@@

Perceptions of inequality have been associated to attitudes about market justice[@lindh_public_2015; @kluegel_social_1995a]. Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].

@@

Among the different approaches to measuring perceived inequality, one of the most widely used is the estimation of wage gaps between occupational extremes, such as between a CEO and a manual worker. This type of item provides a concrete frame that respondents can relate to more easily than abstract questions about national income distribution [@castillo_percepcion_2019; @easterbrook_social_2021; Willis et al., 2015]. While it enables the estimation of inequality using simple heuristics, this method is not without challenges. For instance, people often lack reliable knowledge about the earnings of those at the top of the income ladder, which leads to high variability in responses and the use of biased mental shortcuts [@knell_perceptions_2020]. Despite this, perceived wage gaps are strong predictors of political attitudes[@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.

Another relevant critique of this approach lies in its conflation of different psychological constructs. Many surveys assess perceived inequality through Likert-type items that ask respondents to agree or disagree with statements, such as "income differences are too large," which captures general concern or discomfort rather than a specific perception [@Castillo2011; @garcia-sanchez_vicious_2019].These items mix cognitive estimations with affective evaluations, complicating the interpretation of what respondents perceive versus what they morally reject. As a result, the conceptual clarity between perceived inequality and inequality aversion remains blurred in many empirical studies. To address this limitation, recent work has emphasized the need to distinguish between absolute and comparative measures, as well as between ideal and actual estimates of economic gaps [@garcia-sanchez_creencias_2022; Auspurg et al., 2017].

@@

These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they personally endorse meritocratic principles. This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice (Tejero-Peregrina et al., 2025). Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity. Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_percepcion_2019; @castillo_socialization_2024].

In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups. Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor (Hoyt & Burnette, 2023). This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds. Moreover, Busemeyer, Abrassart, and Nezi (2019) argue that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures. This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.

>>>>

Empirical studies have developed different strategies to capture and operationalize market justice preferences. A common approach in the literature is to gauge attitudes toward the legitimacy of inequality in specific domains, especially when linked to income differences. Kluegel and Smith [-@kluegel_beliefs_1981] and later Kluegel et al. [-@kluegel_legitimation_1999] laid the groundwork for measuring beliefs in market justice through survey items that assess whether people support economic inequality when it reflects personal effort and skills. Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, and/or pensions. For example, @vondemknesebeck_are_2016 and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia. These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts. Some research, like that of Lindh [-@lindh_public_2015] and Svallfors [-@svallfors_political_2007], uses these indicators in combination to measure broader preferences for market-based distribution in welfare systems. More recently, Castillo et al. [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile. These different operationalizations all aim to capture the extent to which individuals accept inequality when it is framed as a reflection of market outcomes.

@@

Perceptions of inequality have been associated to attitudes about market justice [@lindh_public_2015; @kluegel_social_1995a]. Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].

@@

Among the different approaches to measuring perceived inequality, one of the most widely used is the estimation of wage gaps between occupational extremes, such as between a CEO and a manual worker. This type of item provides a concrete frame that respondents can relate to more easily than abstract questions about national income distribution [@castillo_percepcion_2019; @easterbrook_social_2021; @willis_legitimacy_2015]. While it enables the estimation of inequality using simple heuristics, this method is not without challenges. For instance, people often lack reliable knowledge about the earnings of those at the top of the income ladder, which leads to high variability in responses and the use of biased mental shortcuts [@knell_perceptions_2020]. Despite this, perceived wage gaps are strong predictors of political attitudes [@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.

Another relevant critique of this approach lies in its conflation of different psychological constructs. Many surveys assess perceived inequality through Likert-type items that ask respondents to agree or disagree with statements, such as "income differences are too large," which captures general concern or discomfort rather than a specific perception [@Castillo2011; @garcia-sanchez_vicious_2019]. These items mix cognitive estimations with affective evaluations, complicating the interpretation of what respondents perceive versus what they morally reject. As a result, the conceptual clarity between perceived inequality and inequality aversion remains blurred in many empirical studies. To address this limitation, recent work has emphasized the need to distinguish between absolute and comparative measures, as well as between ideal and actual estimates of economic gaps [@garcia-sanchez_creencias_2022; @auspurg_why_2017].

@@

These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they personally endorse meritocratic principles. This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025] Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity. Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_percepcion_2019; @castillo_socialization_2024].

In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups. Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023]. This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds. Moreover, @busemeyer_positive_2021 argue that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures. This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.
